\chapter*{Acknowledgements}

First of all, I would like to acknowledge all the help and guidance I received from my advisor, Dr. Jerry Seitzman, during my time at Georgia Tech.
His passion for research, his critical thinking and his ability to switch focus instantly between macro- and micro-level issues were a constant source of motivation to me.
I am particularly grateful to him for both providing me with the opportunity and freedom to learn so many new things---relevant and sometimes irrelevant to my research---while keeping me focused on the big picture.

I also acknowledge the support and guidance I received from other members of the faculty, particularly Dr. Tim Lieuwen whose goal- and results-driven work ethic helped shape my progress through my Ph.D.
I also want to thank the other members of my committee for giving me valuable inputs and suggestions to improve this thesis.

I would like to thank Bobby, Brad, Sasha, and the rest of the research engineers in the Combustion Lab for helping me at various stages of my research.
I want to acknowledge the invaluable help of Ben Emerson and Luke Walker who assisted me on the LSB project, Kori Rahman who worked with me on the high pressure Raman scattering project and Kyung-Hak Choo who assisted with me on the CH PLIF project.
Special thanks go to Andrew Marshall and Prabhakar Venkateswaran and their undergraduate army for their invaluable work in setting up an LSB flame for me to image with the CH PLIF system.

I am indebted to Mohan, Priya, and others of the ``old gang'' who taught me so much about experimental combustion research during the early days of my graduate life.
I also want to thank Yogi, Yash, Arun, and all the other students at the Combustion Lab for their help and for countless hours of productive discussions.
Last, but not least, I would like to thank my parents for supporting me in all my decisions.

